---
title: "COVID-19 Data Analysis"
author: "Anonymous"
date: "2025-04-20"
output:
  pdf_document: default
---

## Objective

I will use the following data sets from Johns Hopkins University to analyze data pertaining to the COVID-19 pandemic. My objective is to answer the questions below, is well as using an ARIMA model to predict COVID-19 deaths in the US during the first quarter of 2023. 1. Which US state's population was most affected by the COVID-19 virus? 2. How did the United State's mortality rate compare to the rest of the world?

## Data Overview

First, I will import the necessary libraries and import the COVID19 and population data from the five JHU csv files.

```{r import_libraries, message = 'hide'}
library("tidyverse")
library("dplyr")
library("lubridate")
library("forecast")
library("tseries")

```

```{r get_jhu_data}

#import JHU csv files
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",
                "time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_US.csv",
                "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in, file_names)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
```

Now I will read in the data and take an initial look at it.

```{r import_data, message=FALSE}

#read in JHU csv files
us_cases <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
us_deaths <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])
global_population <- read_csv(uid_lookup_url)

#view data
us_cases
global_cases
us_deaths
global_deaths
global_population
```

The COVID data sets contain variables such as province/state, country/region, along with other identifiers that do not initially appear to helpful to my analysis. Each date has a corresponding column, which I will pivot into rows to make more time-series friendly. The population data has one row per city/county, represented by the 'Combined_Key' column.

## Tidy and Transform Data

After taking an initial look at the data, it is evident that some tidying and transformation needs to be done. To start, I will tidy up the global data by pivoting the case and death data and removing some unnecessary columns. I will also filter out all rows where cases are not more than zero, and ensure there is a date column in the correct format.

```{r tidy_global_data}
#pivot and remove unnecessary columns
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region',
                         Lat,
                         Long),
               names_to = "date",
               values_to = "cases") %>% 
  select(-c(Lat, Long))

#pivot and remove unnecessary columns
global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region',
                         Lat,
                         Long),
               names_to = "date",
               values_to = "deaths") %>% 
  select(-c(Lat, Long))

#join both data sets
global <- global_cases %>%
  full_join(global_deaths) %>%
  filter(cases > 0) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

#view new data set
global
```

The same transformations will be applied to the US data.

```{r tidy_us_data, message = FALSE}
#pivot and remove unnecessary columns
us_cases <- us_cases %>%
  pivot_longer(cols = -c(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

#pivot and remove unnecessary columns
us_deaths <- us_deaths %>%
  pivot_longer(cols = -c(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

#join both data sets
us <- us_cases %>%
  full_join(us_deaths) %>%
  filter(cases > 0) %>%
  rename(County = "Admin2")

#view new data set
us
```

In order to truly compare the US data to the global data, I need to create a mutual column and bring populations into the global data set. The column will be called "Combined_Key", which mimics the same column in the US data.

```{r global_pop}
#create a final global data set by performing a left join to only bring in relevant population data
global <- global %>%
  left_join(global_population, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population)

#create combined_key by combining province_state and country_region  
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
```

Now that the tidying and transformations are complete, these final data sets can be used for my analysis.

```{r view_data}
us
summary(us)

global
summary(global)
```

## Exploratory Data Analysis

### Objective #1

For my first objective of determining which US state was most affected by COVID-19, I will summarize cases, deaths, and population by each state and again by the total United States. I will also create variables for cases per million, deaths per million, and mortality rate.

```{r state_summary, message = FALSE}
#Get total state population
state_pop <- us %>%
  distinct(Province_State, County, .keep_all = TRUE) %>%
  group_by(Province_State) %>%
  summarize(Population = sum(Population))

#Aggregate cases/deaths
us_by_state <- us %>%
  group_by(Country_Region, Province_State, date) %>%
  summarize(cases = sum(cases),
            deaths = sum(deaths)) %>%
  ungroup() %>%
  
  #bring in population data
  left_join(state_pop, by = "Province_State") %>%
  filter(Population > 0) %>%
  filter(!is.na(Population)) %>%
  
  #Create new variables
  mutate(deaths_per_mill = deaths * 1000000 / Population,
         cases_per_mill = cases * 1000000 / Population,
         mortality_rate = deaths/ cases) %>%
  select(Province_State, date, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)

#one row per state with accurate totals  
us_states_ovr <- us_by_state %>%
  group_by(Province_State) %>%
  filter(date == max(date)) %>%
  ungroup() %>%
  select(Province_State, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)
  
us_by_state
us_states_ovr
```

Now I will plot my Death per Million variable to identify the top 10 states that were most affected by the COVID-19 deaths.

```{r state_mort_rates}
top_10_states <- us_states_ovr %>%
  arrange(desc(deaths_per_mill)) %>%
  head(10)

ggplot(top_10_states, aes(x = deaths_per_mill, y = reorder(Province_State, deaths_per_mill))) +
  geom_point(color = "darkred", size = 3) +
  labs(title = "COVID-19 Deaths per Million by US State/Province",
       subtitle = "Top 10",
       x = "Deaths per Million",
       y = "State/Province") +
  theme_minimal()
```

The plot shows that relative to population, Arizona was the state most affected by COVID-19 deaths.

### Objective #2

For my second objective of determining how the US's mortality rate compares to the rest of the world, I will now perform the same summarizations and create the same variables, but instead grouping on a national level. I will have 2 data-frames, one containing time-series data and another with a cumulative total.

```{r us_summary, message = FALSE}
#Get total US population
us_pop <- us %>%
  distinct(Country_Region, Province_State, County, .keep_all = TRUE) %>%
  group_by(Country_Region) %>%
  summarize(Population = sum(Population))

#Aggregate cases/deaths
us_totals <- us %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases),
            deaths = sum(deaths)) %>%
  ungroup() %>%
  
  #bring in population data
  left_join(us_pop, by = "Country_Region") %>%
  filter(Population > 0) %>%
  filter(!is.na(Population)) %>%
  
  #Create new variables
  mutate(deaths_per_mill = deaths * 1000000 / Population,
         cases_per_mill = cases * 1000000 / Population,
         mortality_rate = deaths/ cases) %>%
  select(Country_Region, date, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)

#one row per state with accurate totals 
us_ovr <- us_totals %>%
  group_by(Country_Region) %>%
  filter(date == max(date)) %>%
  ungroup() %>%
  select(Country_Region, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)

us_totals
us_ovr
```

The same data-frames will now be built using the global data.

```{r global_summary, message = FALSE}
#Get global populations
global_pop <- global %>%
  distinct(Country_Region, Province_State, .keep_all = TRUE) %>%
  group_by(Country_Region) %>%
  summarize(Population = sum(Population))

#Aggregate cases/deaths globally
global_totals <- global %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases),
            deaths = sum(deaths)) %>%
  ungroup() %>%
  
  #bring in population data
  left_join(global_pop, by = "Country_Region") %>%
  filter(Population > 0) %>%
  filter(!is.na(Population)) %>%
  
  #Create new variables
  mutate(deaths_per_mill = deaths * 1000000 / Population,
         cases_per_mill = cases * 1000000 / Population,
         mortality_rate = deaths/ cases) %>%
  select(Country_Region, date, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)

#one row per country with accurate totals
global_ovr <- global_totals %>%
  group_by(Country_Region) %>%
  filter(date == max(date),
         #Filter out North Korea
         cases > 1) %>%
  ungroup() %>%
  select(Country_Region, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)

global_totals
global_ovr
```

Now that my data-frames are complete, I will merge them together so that the data can be plotted. Since there is a large number of different countries in this data, I will be using a density plot to compare the global COVID-19 mortality rates.

```{r global_mort_rates, warning = FALSE}
#append US summary
merged_data <- bind_rows(global_ovr, us_ovr)

#Density plot
ggplot(merged_data, aes(x = mortality_rate)) +
  geom_density(fill = "lightblue", alpha = 0.5) +
  geom_vline(data = subset(merged_data, Country_Region == "US"),
             aes(xintercept = mortality_rate),
             color = "red", size = 1.2, linetype = "dashed") +
  annotate("text",
           x = subset(merged_data, Country_Region == "US")$mortality_rate,
           y = Inf,
           label = "USA",
           vjust = 8,
           hjust = -.5,
           color = "red") +
  labs(title = "Global COVID-19 Mortality Rate Distribution",
       x = "Mortality Rate",
       y = "Density") +
  scale_x_log10() +
  theme_minimal()
```

The density plot shows that the US has a COVID-19 mortality rate slightly above 1%, which appears to be in line with the global average rate.

### Objective 3

For my third and final objective, I will feed the 'US Totals' data-frame into an ARIMA model to predict COVID-19 deaths during the first quarter of 2023. The model will be trained using the data from 2020-2022, and the predicted deaths will be compared to the actual deaths for the first quarter of 2023.

```{r model_deaths}
#Filter out days with 0 deaths
model_data <- us_totals %>%
  filter(deaths > 0) %>%
  select(date, deaths)

# Split data into training (2020-2022) and testing (2023)
train_data <- model_data %>% filter(date < as.Date("2023-01-01"))
test_data <- model_data %>% filter(date >= as.Date("2023-01-01"))


# Convert training data to time series object
ts_train <- ts(train_data$deaths, start = c(2020, 1), frequency = 365)

# Convert testing data to time series object
ts_test <- ts(test_data$deaths, start = c(2023, 1), frequency = 365)

# Apply 2nd Differencing on training data for stationarity
diff_train <- diff(diff(ts_train))

# p-value is greater than .05, stationarity achieved
adf.test(diff_train)

# Fit ARIMA model to training data only
arima_model <- auto.arima(diff_train)
summary(arima_model)

# Forecast for the length of the testing set
forecasted <- forecast(arima_model, h = length(ts_test))

# Reverse differencing for testing period
forecasted_differences <- as.numeric(forecasted$mean)
first_cumsum <- cumsum(forecasted_differences) + as.numeric(tail(diff(ts_train), n = 1))
original_scale_predictions <- cumsum(first_cumsum) + as.numeric(tail(ts_train, n = 1))

# Create results data frame for predictions
predicted_dates <- seq(
  from = as.Date("2023-01-01"),
  by = "day",
  length.out = length(original_scale_predictions)
)

#Actual deaths for 2020-2022
actual_deaths <- model_data %>%
  filter(date <= as.Date("2022-12-31"))

#Combine actual and predicted deaths
comparison <- bind_rows(
  train_data %>% filter(date >= as.Date("2023-01-01")),  # Keep actual deaths for 2023
  data.frame(date = predicted_dates, deaths = test_data$deaths, predicted_deaths = original_scale_predictions)) %>%
  filter(year(date) == 2023)  # Filter to include only 2023 data

#Plot predicted vs actual deaths
ggplot(comparison, aes(x = date)) +
  geom_line(aes(y = deaths, color = "Actual"), size = 1, na.rm = TRUE) +
  geom_line(aes(y = predicted_deaths, color = "Predicted"), size = 1, na.rm = TRUE) +
  scale_color_manual(values = c("Actual" = "black", "Predicted" = "red")) +
  labs(
    title = "Actual vs Predicted COVID-19 Deaths",
    subtitle = "Using ARIMA with Differencing (Trained on 2020-2022, Tested on 2023)",
    x = "Date",
    y = "Daily Deaths",
    color = "Legend") +
  theme_minimal()
```

## Conclusion

By using the data sets from Johns Hopkins University, I was able to complete all my objectives. However, it is important to discuss potential biases in my analysis. There are many factors that influence COVID-19 cases, deaths, and by association, mortality rates. the data sets provided do not account for variables such as government policy, vaccine rates, or the time between diagnosis and death (lag). In the United States, these variables would be different across both states and cities. Globally, many countries had very strict COVID-19 policies, while many countries did not have much policy at all. When interpreting the results of my analysis, it is important to remember that the data does not account for these types of variables, making my findings more exploratory than factual.
